---
title: "stan_first"
output: html_document
date: "2026-02-27"
---

#might need to change the simulation to match the model more (currently noise-based, with no WSLS values)
#need to add WSLS probs (win_stay, lose_shift)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

#load packages
```{r}
pacman::p_load(
  tidyverse, # Data manipulation and plotting
  dplyr,
  here,         # Robust file paths
  cmdstanr,     # Stan interface
  posterior,    # Posterior analysis
  bayesplot,    # Plotting MCMC
  furrr,        # Parallel processing
  future        # Future backend
)

# Parallel Setup: Leave 2 cores free for system stability
n_cores <- parallel::detectCores() - 2
plan(multisession, workers = n_cores)

# Optional: Set a default ggplot theme
theme_set(theme_classic())
```

#create the data (vars, basic model logic)
```{r}
trials <- 120 # Number of trials per simulated agent

# Define the agent function (copied from previous chapter)
WSLSAgent_noise_f <- function(prevChoice, feedback, noise, win_stay = 0.7, lose_shift = 0.9){
  #win_stay and lose_shift
  if (feedback == 1 && runif(1) < win_stay) {
    choice = prevChoice
  } else {
    choice <- ifelse(runif(1) < lose_shift, 1 - prevChoice, prevChoice) #if not win, do this
    }
  
  if (noise > 0 && rbinom(1, 1, noise) == 1) { #if noise, override with 50/50
    choice <- rbinom(1, 1, 0.5)
  }
  return(choice)
}

RandomAgent_f <- function(rate){
  choice <- rbinom(1, 1, rate)
  return(choice)
}
```

#create the data 2 (WSLS function for simulation)
```{r}
#simulate the wsls agent separately (to have the feedback and such)
Simulate_WSLS_agent_noise_f <- function(noise, opponent, trials = 120) {
  choices  <- integer(trials)
  feedback <- integer(trials)
  
  #first is random
  choices[1] <- rbinom(1, 1, 0.5)

  for (t in 2:trials) {
    feedback[t - 1] <- as.integer(opponent[t - 1] == choices[t - 1])
    choices[t] <- WSLSAgent_noise_f(choices[t - 1], feedback[t - 1], noise)
  }
  # Fill in the last trial's feedback
  feedback[trials] <- as.integer(opponent[trials] == choices[trials])
  
  return(list(choices = choices, feedback = feedback))
}
```

#create the data 3 (simulate it)
```{r}
sim_data_dir <- "simdata"
sim_data_file <- here(sim_data_dir, "04_WSLS_noise.csv")

# Flag to control whether to regenerate simulation/fitting results for the knitted files
# Set to TRUE to rerun everything (takes time!), FALSE to load saved results
regenerate_simulations <- TRUE

if (regenerate_simulations || !file.exists(sim_data_file)) {
  cat("Generating new simulation data...\n")
  # Use expand_grid for cleaner parameter combinations
  param_grid <- expand_grid(
    noise = seq(0, 0.5, 0.1) # Noise levels
  )
  
  #WSLS needs an opponent (random, 0.5)
  opponent <- map_int(1:trials, ~RandomAgent_f(rate = 0.5))
  cat(sum(opponent), " this is how many times the opponent chose right!\n")
  
  # Use map_dfr for efficient simulation across parameters
  d <- pmap_dfr(param_grid, function(noise) {
  # Simulate choices for one agent/condition (do in full since need feedback)
  result <- Simulate_WSLS_agent_noise_f(noise, opponent, trials)
  
  # Create tibble for this condition
  tibble(
    trial            = 1:trials,
    WSLS_choice      = result$choices,
    feedback         = result$feedback,
    true_noise       = noise,
    opponent_choices = opponent,
  )
}) %>%
  # Calculate cumulative rate for visualization
  group_by(true_noise) %>%
  mutate(cumulative_rate_WSLS = cumsum(WSLS_choice) / row_number()) %>% #cumulative rat of choosing right
  mutate(cumulative_rate_opponent= cumsum(opponent_choices) / row_number()) %>%
  ungroup()

  # Save the generated data
  write_csv(d, sim_data_file)
  cat("Generated new simulation data and saved to", sim_data_file, "\n")
} else {
  # Load existing simulation data
  d <- read_csv(sim_data_file)
  cat("Loaded existing simulation data from", sim_data_file, "\n")
}
```

#Putting a subset of data into a string so it can go into stan, need to do this with everything
```{r}
# Subset data for one specific agent simulation (rate=0.8, noise=0)
d1 <- d %>% filter(true_noise == 0.0) # Make sure noise level matches

data_for_stan <- list(
  n        = trials, # Number of trials (integer)
  h        = d1$WSLS_choice, # Vector of choices (0s and 1s) for this agent
  feedback = d1$feedback      #feedback
)

str(data_for_stan) # Show the structure of the list

#print(d)
```

#visualize the data
```{r}
# --- Visualize Simulated Data ---
# Plot cumulative rate, faceted by noise level
p1 <- ggplot(d, aes(x = trial, y = cumulative_rate_WSLS, color = "Rate WSLS")) +
  geom_line(alpha = 0.8) +
  geom_line(aes(y=cumulative_rate_opponent, color = "Rate Opponent (Random 0.5)")) + 
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  scale_color_viridis_d() + # Use perceptually uniform color scale (non continuous)
  facet_wrap(~paste("Noise =", true_noise)) + # Facet by noise level
  labs(
    title = "Simulated Choice Behavior of WSLS (Cumulative Rate)",
    subtitle = "WSLS Agents with different bias rates and noise levels",
    x = "Trial Number",
    y = "Cumulative Proportion Choosing 'Right'",
    color = "Players"
  ) +
  theme_minimal() +
  ylim(0, 1)

print(p1)
```


#call the model
```{r}
## --- Compile and Fit ---
# Specify model file path
stan_file_bernoulli <- here("stan_model_actual_WSLS.stan")
# Define path for saving fitted model object
stan_results_dir <- "simmodels"
model_file_bernoulli <- here(stan_results_dir, "stan_model_actual_WSLS.stan")

if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)

if (regenerate_simulations || !file.exists(model_file_bernoulli)) {
  cat("Compiling and fitting Stan model (Bernoulli)...\n")
  # Compile the Stan model (only needs to be done once unless code changes)
  mod_bernoulli <- cmdstan_model(stan_file_bernoulli,
                                 cpp_options = list(stan_threads = FALSE), # Enable threading
                                 stanc_options = list("O1")) # Basic optimization

  # Sample from the posterior distribution using MCMC
  fit_bernoulli <- mod_bernoulli$sample(
    data = data_for_stan,          # The data list we prepared
    seed = 123,                    # For reproducible MCMC sampling
    chains = 4,                    # Number of parallel Markov chains (recommend 4)
    parallel_chains = 4, # Run chains in parallel
    #threads_per_chain = 1,         # For within-chain parallelization (usually 1 is fine)
    iter_warmup = 1000,            # Number of warmup iterations per chain (discarded)
    iter_sampling = 2000,          # Number of sampling iterations per chain (kept)
    refresh = 0,                   # How often to print progress updates. 0 means never since the model is so fast
    max_treedepth = 10,            # Controls complexity of MCMC steps (adjust if needed)
    adapt_delta = 0.8              # Target acceptance rate (adjust if divergences occur)
  )

  # Save the fitted model object
  fit_bernoulli$save_object(file = model_file_bernoulli)
  cat("Model fit completed and saved to:", model_file_bernoulli, "\n")
} else {
  # Load existing fitted model object
  fit_bernoulli <- readRDS(model_file_bernoulli)
  cat("Loaded existing model fit from:", model_file_bernoulli, "\n")
}


```

#Health checks of the model
```{r}
# Extract the summary and diagnostics from our fitted model object
fit_summary <- fit_bernoulli$summary()
diagnostic_summary <- fit_bernoulli$diagnostic_summary()

print(diagnostic_summary)
```

```{r}
# 1. Check for Divergent Transitions (Zero Tolerance)
divergences <- sum(diagnostic_summary$num_divergent)
if (divergences > 0) {
  stop(paste("CRITICAL FAILURE:", divergences, "divergences detected. The posterior geometry is pathological. Reparameterization required. Do not interpret these results."))
}

print(divergences)
```

```{r}
# 2. Check E-BFMI (Energy-Bayesian Fraction of Missing Information)
# Threshold is typically 0.3. Values below this indicate poor exploration.
min_ebfmi <- min(diagnostic_summary$ebfmi)
if (min_ebfmi < 0.3) {
  warning(paste("WARNING: Low E-BFMI detected (Min =", round(min_ebfmi, 3), 
                "). The sampler may struggle to explore the tails of the posterior."))
}

print(min_ebfmi)
```

```{r}
# 3. Check for Convergence (Rhat < 1.01)
max_rhat <- max(fit_summary$rhat, na.rm = TRUE)
if (max_rhat > 1.01) {
  stop(paste("CRITICAL FAILURE: Chains did not converge. Max Rhat =", round(max_rhat, 3)))
}

print(max_rhat)
```

```{r}
# 4. Check Effective Sample Size (ESS > 400 for both bulk and tail)
min_ess_bulk <- min(fit_summary$ess_bulk, na.rm = TRUE)
min_ess_tail <- min(fit_summary$ess_tail, na.rm = TRUE)
if (min_ess_bulk < 400 || min_ess_tail < 400) {
  warning("WARNING: Insufficient Effective Sample Size. Autocorrelation is too high. Run longer chains.")
}

print(min_ess_bulk)
print(min_ess_tail)
```

#prior predictive checks
```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
h_prior_rep  <- fit_bernoulli$draws("prior_preds", format = "matrix")
prior_rep_sums <- rowSums(h_prior_rep) #amount of rights
observed_sum <- sum(data_for_stan$h)

ppc_prior_ggplot <- ggplot(tibble(Replicated_Sum = prior_rep_sums), 
                           aes(x = Replicated_Sum)) +
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  geom_vline(xintercept = observed_sum, color = "darkred", linewidth = 1.2) +
  ggtitle("Prior Predictive Check: Total 'Right' Choices") +
  xlab(paste("Total Number of 'Right' Choices (out of", trials, "trials)")) +
  ylab("Frequency") +
  theme_classic()

ppc_prior_ggplot
#note: we only have one observed sum
```

#posterior predective checks 
```{r}
# Extract posterior predictive simulations from the fitted object
# This matrix has dimensions: [number_of_draws, number_of_trials]
h_post_rep <- fit_bernoulli$draws("posterior_preds", format = "matrix")

# Calculate the test statistic T(y) for the *actual* empirical data
observed_sum <- sum(data_for_stan$h) 

# Calculate the test statistic T(y_rep) for *each* replicated dataset
# We compute the row-wise sum across the posterior draws
post_rep_sums <- rowSums(h_post_rep)

# Bundle into a tibble for strictly structured plotting
ppc_df <- tibble(
  Replicated_Sum = post_rep_sums
)

# --- Graphical Posterior Predictive Check using ggplot2 ---
ppc_posterior_ggplot <- ggplot(ppc_df, aes(x = Replicated_Sum)) +
  geom_histogram(fill = "skyblue", color = "white", alpha = 0.7, bins = 30) +
  geom_vline(xintercept = observed_sum, color = "black", linetype = "dashed", linewidth = 1) +
  ggtitle("Posterior Predictive Check: Total 'Right' Choices") +
  xlab(paste("Total Number of 'Right' Choices (out of", trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_sum, y = Inf, 
    label = "Empirical data", 
    vjust = 2, hjust = if(observed_sum > trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic()

ppc_posterior_ggplot
```
```{r}
ppc_overlay_df <- tibble(
  Replicated_Sum = c(prior_rep_sums, post_rep_sums),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(post_rep_sums)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ppc_overlay_ggplot <- ggplot(ppc_overlay_df, aes(x = Replicated_Sum, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = observed_sum, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Generative Evolution: Prior vs. Posterior Predictive Checks") +
  xlab(paste("Total Number of 'Right' Choices (out of", trials, "trials)")) +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_sum, y = Inf, 
    label = "Empirical data", 
    vjust = 2, hjust = if(observed_sum > trials/2) 1.1 else -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(ppc_overlay_ggplot)
#extremes because WSLS tends to repeat choices, so a lot of left choices, or a lot of right choices
```

```{r}
# --- Example of checking another statistic (e.g., max run length) ---
# Define a function to calculate the maximum run length in a vector
max_run_length <- function(x) {
  if (length(x) == 0 || all(is.na(x))) return(0) 
  rle_x <- rle(x) # Run Length Encoding calculates consecutive identical choices
  if (length(rle_x$lengths) == 0) return(0) 
  max(rle_x$lengths) 
}

# 2. Calculate the empirical test statistic T(y)
observed_max_run <- max_run_length(data_for_stan$h)

# 3. Extract prior replicated matrices and compute T(y_rep) for the prior
# h_prior_rep is an S x N matrix generated from the prior distributions
#h_prior_rep <- fit_gq$draws("h_prior_rep", format = "matrix")
h_prior_rep <- fit_bernoulli$draws("prior_preds", format = "matrix")
prior_max_runs <- apply(h_prior_rep, 1, max_run_length)

# 4. Extract posterior replicated matrices and compute T(y_rep) for the posterior
# h_post_rep is an S x N matrix generated from the posterior distributions
#h_post_rep <- fit_gq$draws("h_post_rep", format = "matrix") #what is fit_gq??
h_post_rep <- fit_bernoulli$draws("posterior_preds", format = "matrix")
post_max_runs <- apply(h_post_rep, 1, max_run_length)

# 5. Bundle into a strictly structured tibble for overlay plotting
ppc_run_df <- tibble(
  Replicated_Max_Run = c(prior_max_runs, post_max_runs),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(post_max_runs)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 6. Generate the overlay plot
p_ppc_max_run <- ggplot(ppc_run_df, aes(x = Replicated_Max_Run, fill = Distribution)) +
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 20) +
  geom_vline(xintercept = observed_max_run, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Generative Evolution: Prior vs. Posterior (Maximum Run Length)") +
  xlab("Test Statistic T(y_rep): Maximum Run Length of Consecutive Choices") +
  ylab("Frequency (Replicated Datasets)") +
  # Add annotation for the empirical ground truth
  annotate(
    "text", x = observed_max_run, y = Inf, 
    label = "Empirical T(y)", 
    vjust = 2, hjust = -0.1, 
    size = 4, fontface = "bold"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )

print(p_ppc_max_run)
#has learned?
```
#More prior & posterior checks (now for each val, for win_stay and lose_shift)

#WinSTAY
```{r}
#prior & posteriors again
ws_prior <- fit_bernoulli$draws("ws_prior", format = "matrix")
ws_posterior <- fit_bernoulli$draws("win_stay", format = "matrix")
#get the real winstay (in this case known, and only one val)
win_stay_real = 0.7

ppc_winstay <- tibble(
  Win_stay = c(ws_prior, ws_posterior),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(ws_prior)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ggplot(ppc_winstay, aes(x = Win_stay, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = win_stay_real, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Win_Stay: Prior vs. Posterior Predictive Checks") +
  xlab(paste("Win_stay prior & posterior compared")) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  ) +
  annotate(
    "text",
    x = win_stay_real,
    y = Inf,
    label = "Empirical Win_stay",
    vjust = 2,
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  )
```
#loseShift
```{r}
#prior & posteriors again
ls_prior <- fit_bernoulli$draws("ls_prior", format = "matrix")
ls_posterior <- fit_bernoulli$draws("lose_shift", format = "matrix")
#get the real winstay (in this case known, and only one val)
lose_shift_real = 0.9

ppc_loseshift <- tibble(
  Lose_shift = c(ls_prior, ls_posterior),
  Distribution = factor(
    rep(c("Prior Predictive", "Posterior Predictive"), each = length(ls_prior)),
    levels = c("Prior Predictive", "Posterior Predictive")
  )
)

# 2. Generate the overlay plot
ggplot(ppc_loseshift, aes(x = Lose_shift, fill = Distribution)) +
  # Use position = "identity" and alpha to ensure both distributions are visible
  geom_histogram(position = "identity", alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = lose_shift_real, color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Prior Predictive" = "#ffbaba", "Posterior Predictive" = "skyblue")) +
  ggtitle("Lose_shift: Prior vs. Posterior Predictive Checks") +
  xlab(paste("Lose_shift prior & posterior compared")) +
  theme_classic() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  ) +
  annotate(
    "text",
    x = lose_shift_real-0.3,
    y = Inf,
    label = "Empirical Lose_shift",
    vjust = 2,
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  )
```
```
#it overestimates winstay & underestimates loseshift: this is because of noise & the values "trading off"????
#but why are these plots... identical??? (maybe recheck the model? or the simulation? or the plots?)
